---
title: "Data Visualization"
author: "Introduction to R for Public Health Researchers"
output:
  beamer_presentation: default
  ioslides_presentation:
    css: ../../styles.css
    widescreen: yes
---

```{r knit-setup, echo=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, 
               message = FALSE, 
               warning = FALSE,
               fig.height = 3.5,
               fig.width = 3.5, 
               comment = "")
```

## Basic Plots

We covered some basic plots previously, but we are going to expand the ability to customize these basic graphics first. 

```{r seed, comment="",echo=FALSE,prompt=TRUE, }
set.seed(3) 
```


## Read in Data

```{r plotEx,  fig.align='center',cache=FALSE}
library(readr)
death = read_csv(
  "http://www.aejaffe.com/winterR_2017/data/indicatordeadkids35.csv")
death[1:2, 1:5]
```


```{r}
colnames(death)[1] = "country"
head(death)
```


## Basic Plots

```{r plot1, comment="",prompt=TRUE,  fig.align='center',cache = FALSE}
library(dplyr)
sweden = death %>% 
  filter(country == "Sweden") %>% 
  select(-country)
year = as.numeric(colnames(sweden))
plot(as.numeric(sweden) ~ year)
```


## Basic Plots

The y-axis label isn't informative, and we can change the label of the y-axis using `ylab` (`xlab` for x), and `main` for the main title/label.

```{r plotEx2, comment="",prompt=TRUE,  fig.align='center',cache = FALSE}
plot(as.numeric(sweden) ~ year,
      ylab = "# of deaths per family", main = "Sweden", type = "l")
```


## Basic Plots
Let's drop any of the projections and keep it to year 2012, and change the points to blue. 
```{r plotEx3, fig.align='center', cache = FALSE}
plot(as.numeric(sweden) ~ year,
      ylab = "# of deaths per family", main = "Sweden",
     xlim = c(1760,2012), pch = 19, cex=1.2,col="blue")
```


## Basic Plots
You can also use the `subset` argument in the `plot()` function, only when using formula notation:

```{r plotEx_sub, fig.align='center', cache = FALSE}
plot(as.numeric(sweden) ~ year,
      ylab = "# of deaths per family", main = "Sweden",
     subset = year < 2015, pch = 19, cex=1.2,col="blue")
```

## Reshape the data

After reshaping the data to long, we can plot the data with one `data.frame`:

```{r makelong_swede, fig.align='center', cache = FALSE}
library(tidyr)
long = gather(death, key = year, value = deaths, -country)
long = long %>% filter(!is.na(deaths))
head(long)
class(long$year)
long$year = as.numeric(long$year)
```

## Plot the long data

```{r plot_long_swede, fig.align='center'}
swede_long = long %>% filter(country == "Sweden")
plot(deaths ~ year, data = swede_long)
```

## ggplot2

`ggplot2` is a package of plotting that is very popular and powerful (using the **g**rammar of **g**raphics). `qplot` is a short hand for "quick plot", which can usually be similar to `plot`


```{r}
library(ggplot2)
qplot(x = year, y = deaths, data = swede_long)
```


## ggplot2

The generic plotting function is `ggplot`, which uses **aes**thetics:

```
ggplot(data, aes(args))
```

```{r generic_gg, comment="", prompt=TRUE, fig.align='center', cache=FALSE}
g = ggplot(data = swede_long, aes(x = year, y = deaths))
```

`g` is an object, which you can adapt into multiple plots!


## ggplot2

Common aesthetics:

- **x**
- **y**
- **colour/color**
- **size**
- **fill** 
- *shape**


If you set these in `aes`, you set them to a variable.  If you want to set them for all values, set them in a `geom`.


## ggplot2: what's a `geom`?

`g` on it's own can't be plotted, we have to add `layers`, usually with `geom_` commands:

- `geom_point` - add points
- `geom_line` - add lines
- `geom_density` - add a density plot
- `geom_histogram` - add a histogram
- `geom_smooth` - add a smoother
- `geom_boxplot` - add a boxplots
- `geom_tile` - rectangles/heatmaps


## ggplot2: adding a geom and assigning

You "add" things to a plot with a `+` sign (not pipe!).  If you assign a plot to an object, you must call `print` to print it.  

```{r gprint_point}
gpoints = g + geom_point()
print(gpoints)
```

## ggplot2: adding a geom

Otherwise it prints by default

```{r geom_line}
g + geom_line()
```

## ggplot2: adding a geom

You can add multiple geoms:

```{r geom_line_point}
g + geom_line() + geom_point()
```

## ggplot2: grouping - using colour

If we want a plot with new data, we have to call `ggplot` again.  Let's group plots by `country` using `colour`:

```{r geom_all}
g = ggplot(long, aes(x = year, y = deaths, 
  colour = country))
g + geom_line()
```

## ggplot2: grouping - using colour

Let's remove the legend using the `guide` command:

```{r geom_noguide}
g + geom_line() + guides(colour = FALSE)
```

## ggplot2: boxplot

```{r geom_box}
g + geom_boxplot()
```


## Bar Plots
* Stacked Bar Charts are sometimes wanted to show distributions of data

```{r barplot2, fig.align='center', cache = FALSE}
## Stacked Bar Charts
cars = read_csv(
  "http://www.aejaffe.com/winterR_2017/data/kaggleCarAuction.csv")
counts <- table(cars$IsBadBuy, cars$VehicleAge)
barplot(counts, main="Car Distribution by Age and Bad Buy Status",
  xlab="Vehicle Age", col=c("darkblue","red"),
    legend = rownames(counts))
```

## Bar Plots
`prop.table` allows you to convert a table to proportions (depends on margin - either row percent or column percent)

```{r barplot2a, fig.align='center', cache = FALSE}
## Use percentages (column percentages)
barplot(prop.table(counts, 2), main="Car Distribution by Age and Bad Buy Status",
  xlab="Vehicle Age", col=c("darkblue","red"),
    legend = rownames(counts))
```

## Bar Plots

Using the `beside` argument in `barplot`, you can get side-by-side barplots.
```{r barplot3, fig.align='center', cache = FALSE}
# Stacked Bar Plot with Colors and Legend    
barplot(counts, main="Car Distribution by Age and Bad Buy Status",
  xlab="Vehicle Age", col=c("darkblue","red"),
    legend = rownames(counts), beside=TRUE)
```

## Graphics parameters

Set within most plots in the base 'graphics' package:

* pch = point shape, http://voteview.com/symbols_pch.htm
* cex = size/scale
* xlab, ylab = labels for x and y axes
* main = plot title
* lwd = line density
* col = color
* cex.axis, cex.lab, cex.main = scaling/sizing for axes marks, axes labels, and title

## Devices

By default, R displays plots in a separate panel. From there, you can export the plot to a variety of image file types, or copy it to the clipboard. 

However, sometimes its very nice to save many plots made at one time to one pdf file, say, for flipping through. Or being more precise with the plot size in the saved file.

R has 5 additional graphics devices: `bmp()`, `jpeg()`, `png()`, `tiff()`, and `pdf()`

## Devices

The syntax is very similar for all of them:

```
pdf("filename.pdf", width=8, height=8) # inches
plot() # plot 1
plot() # plot 2
# etc
dev.off()
```

Basically, you are creating a pdf file, and telling R to write any subsequent plots to that file. Once you are done, you turn the device off. Note that failing to turn the device off will create a pdf file that is corrupt, that you cannot open.

## Boxplots, revisited

These are one of my favorite plots. They are way more informative than the barchart + antenna...

```{r boxplots, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
boxplot(weight ~ Diet, data=ChickWeight, outline=FALSE)
points(ChickWeight$weight ~ jitter(as.numeric(ChickWeight$Diet),0.5))
```

## Formulas

Formulas have the format of `y ~ x` and functions taking formulas have a `data` argument where you pass the data.frame.  You don't need to use `$` or referencing when using formulas:

```{r box_ex, eval=FALSE}
boxplot(weight ~ Diet, data=ChickWeight, outline=FALSE)
```


## Colors

R relies on color 'palettes'.

```{r pal, fig.align='center', cache = FALSE}
palette("default")
plot(1:8, 1:8, type="n")
text(1:8, 1:8, lab = palette(), col = 1:8)
```


## Colors

The default color palette is pretty bad, so you can try to make your own. 

```{r pal2, fig.align='center', cache = FALSE}
palette(c("darkred","orange","blue"))
plot(1:3,1:3,col=1:3,pch =19,cex=2)
```


## Colors

It's actually pretty hard to make a good color palette. Luckily, smart and artistic people have spent a lot more time thinking about this. The result is the `RColorBrewer` package

`RColorBrewer::display.brewer.all()` will show you all of the palettes available. You can even print it out and keep it next to your monitor for reference.

The help file for `brewer.pal()` gives you an idea how to use the package.

You can also get a "sneak peek" of these palettes at: http://colorbrewer2.org/ . You would provide the number of levels or classes of your data, and then the type of data: sequential, diverging, or qualitative. The names of the `RColorBrewer` palettes are the string after 'pick a color scheme:' 


## Colors


```{r pal3, fig.align='center', cache=FALSE}
palette("default")
plot(weight ~ Time, data= ChickWeight, pch = 19, col = Diet)
```

## Colors

```{r pal4, fig.align='center', cache=FALSE}
library(RColorBrewer)
palette(brewer.pal(5,"Dark2"))
plot(weight ~ Time, data=ChickWeight, pch = 19,  col = Diet)
```


## Colors

```{r pal5, fig.align='center', cache=FALSE}
library(RColorBrewer)
palette(brewer.pal(5,"Dark2"))
plot(weight ~ jitter(Time,amount=0.2),data=ChickWeight,
     pch = 19,  col = Diet,xlab="Time")
```



## Adding legends

The legend() command adds a legend to your plot. There are tons of arguments to pass it. 

x, y=NULL: this just means you can give (x,y) coordinates, or more commonly just give x, as a character string: "top","bottom","topleft","bottomleft","topright","bottomright".

legend: unique character vector, the levels of a factor

pch, lwd: if you want points in the legend, give a pch value. if you want lines, give a lwd value. 

col: give the color for each legend level


## Adding legends


```{r leg1, fig.align='center', cache=FALSE}
palette(brewer.pal(5,"Dark2"))
plot(weight ~ jitter(Time,amount=0.2),data=ChickWeight,
                pch = 19,  col = Diet,xlab="Time")
legend("topleft", paste("Diet",levels(ChickWeight$Diet)), 
        col = 1:length(levels(ChickWeight$Diet)),
       lwd = 3, ncol = 2)
```


## Coloring by variable

```{r circ, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
circ = read_csv("http://www.aejaffe.com/winterR_2017/data/Charm_City_Circulator_Ridership.csv")
palette(brewer.pal(7,"Dark2"))
dd = factor(circ$day)
plot(orangeAverage ~ greenAverage, data=circ, 
     pch=19, col = as.numeric(dd))
legend("bottomright", levels(dd), col=1:length(dd), pch = 19)
```

## Coloring by variable

```{r circ2, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
dd = factor(circ$day, levels=c("Monday","Tuesday","Wednesday",
            "Thursday","Friday","Saturday","Sunday"))
plot(orangeAverage ~ greenAverage, data=circ,
     pch=19, col = as.numeric(dd))
legend("bottomright", levels(dd), col=1:length(dd), pch = 19)
```

## ggplot2

`ggplot2` is a package of plotting that is very popular and powerful. `qplot` is a short hand for "quick plot". We can simply do a boxplot:

```{r geoboxplot, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
library(ggplot2)
qplot(factor(Diet), y = weight, 
      data = ChickWeight, geom = "boxplot")
```

## ggplot2

The generic plotting function is `ggplot`:

```{r geoboxplot_g, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
g = ggplot(aes(x = Diet, y = weight), data = ChickWeight)
g + geom_boxplot()
```



## Boxplots revisited again

We can do the same plot, by just saying we want a boxplot and points (and jitter the points)

```{r geoboxpoint, comment="",prompt=TRUE, fig.align='center', cache = FALSE}
qplot( factor(Diet), y = weight, data = ChickWeight, 
       geom = c("boxplot", "jitter"))
```


## ggplot2: Adding 2 geoms together

To have multiple geometrics, just "add" them

```{r geoboxplot_add, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
g + geom_boxplot() + geom_point(position = "jitter")
```


## ggplot2: Adding 2 geoms together

To have multiple geometrics, just "add" them

```{r geoboxplot_addjitter, fig.align='center', cache=FALSE}
g + geom_boxplot() + geom_jitter()
```

## Histograms again
 
We can do histograms again using `hist`.  Let's do histograms of weight at all time points for the chick's weights.  We reiterate how useful these are to show your data.

```{r hist, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
hist(ChickWeight$weight, breaks = 20)
```

## Multiple Histograms

```{r ghist, comment="",prompt=TRUE, fig.align='center', cache = FALSE}
qplot(x = weight, 
      fill = factor(Diet),
      data = ChickWeight, 
      geom = c("histogram"))
```

## Multiple Histograms

Alpha refers tot he opacity of the color, less is 
```{r ghist_alpha, comment="",prompt=TRUE, fig.align='center', cache = FALSE}
qplot(x = weight, fill = Diet, data = ChickWeight, 
      geom = c("histogram"), alpha=I(.7))
```

## Multiple Densities

We cold also do densities
```{r gdens, comment="",prompt=TRUE, fig.align='center', cache = FALSE}
qplot(x= weight, fill = Diet, data = ChickWeight, 
      geom = c("density"), alpha=I(.7))
```


## Multiple Densities

```{r gdens_alpha, comment="",prompt=TRUE, fig.align='center', cache = FALSE}
qplot(x= weight, colour = Diet, data = ChickWeight, 
      geom = c("density"), alpha=I(.7))
```


## Multiple Densities

```{r gdens_alpha_gg, comment="",prompt=TRUE, fig.align='center', cache = FALSE}
ggplot(aes(x= weight, colour = Diet), 
  data = ChickWeight) + geom_density(alpha=I(.7))
```

## Multiple Densities

You can take off the lines of the bottom like this
```{r gdens_line_alpha, comment="",prompt=TRUE, fig.align='center', cache = FALSE}
ggplot(aes(x = weight, colour = Diet), data = ChickWeight) + 
  geom_line(stat = "density")
```

## Spaghetti plot
 
We can make a spaghetti plot by telling ggplot we want a "line", and each line is colored by Chick.

```{r spaghetti, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
qplot(x=Time, y=weight, colour = Chick, 
      data = ChickWeight, geom = "line")
```



## Spaghetti plot: Facets

In ggplot2, if you want separate plots for something, these are referred to as facets.

```{r fac_spag, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
qplot(x = Time, y = weight, colour = Chick, 
      facets = ~Diet, data = ChickWeight, geom = "line")
```


## Spaghetti plot: Facets

We can turn off the legend (referred to a "guide" in ggplot2). (Note - there is different syntax with the `+`)
```{r fac_spag_noleg, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
qplot(x=Time, y=weight, colour = Chick, 
      facets = ~ Diet,  data = ChickWeight, 
        geom = "line") + guides(colour=FALSE)
```

## Spaghetti plot: Facets

```{r fac_spag2, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
ggplot(aes(x = Time, y = weight, colour = Chick), 
    data = ChickWeight) + geom_line() + 
    facet_wrap(facets = ~Diet) + guides(colour = FALSE)
```


## ggplot2

Let's try this out on the childhood mortality data used above. However, let's do some manipulation first, by using `gather` on the data to convert to long. 

```{r}
library(tidyr)
long = death
long$state = rownames(long)
long = long %>% gather(year, deaths, -state)
head(long, 2)
```

## ggplot2

Let's also make the year numeric, as we did above in the stand-alone `year` variable. 

```{r}
library(stringr)
library(dplyr)
long$year = long$year %>% str_replace("^X", "") %>% as.numeric
long = long %>% filter(!is.na(deaths))
```

## ggplot2

```{r geom_line_qplot, comment="",prompt=TRUE, fig.align='center', cache=FALSE}
qplot(x = year, y = deaths, colour = state, 
    data = long, geom = "line") + guides(colour = FALSE)
```

## ggplot2

Let's try to make it different like base R, a bit.  We use `tile` for the geometric unit:

```{r geom_tile}
qplot(x = year, y = state, colour = deaths, 
    data = long, geom = "tile") + guides(colour = FALSE)
```

## ggplot2 

Useful links:

* http://docs.ggplot2.org/0.9.3/index.html
* http://www.cookbook-r.com/Graphs/

