---
title: "Manipulating Data in R"
author: "John Muschelli"
date: "June 16, 2016"
output:
  beamer_presentation: default
  ioslides_presentation:
    css: ../../styles.css
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(comment = "")
```


## Reshaping Data 

In this module, we will show you how to:

1. Reshaping data from long (tall) to wide (fat)
2. Reshaping data from wide (fat) to long (tall)
3. Merging Data
4. Perform operations by a grouping variable

## Setup

We will show you how to do each operation in base R then show you how to use the `dplyr` or `tidyr` package to do the same operation (if applicable).  

See the "Data Wrangling Cheat Sheet using `dplyr` and `tidyr`":

* https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

## Data used: Charm City Circulator

http://www.aejaffe.com/summerR_2016/data/Charm_City_Circulator_Ridership.csv

```{r}
circ = read.csv("http://www.aejaffe.com/summerR_2016/data/Charm_City_Circulator_Ridership.csv", as.is = TRUE)
head(circ, 2)
```

## Creating a Date class from a character date

```{r, message= FALSE}
library(lubridate) # great for dates!
library(dplyr) # mutate/summarise functions
circ = mutate(circ, date = mdy(date))
sum( is.na(circ$date) ) # all converted correctly
head(circ$date)
class(circ$date)
```

## Making column names a little more separated

We will use `str_replace` from `stringr` to put periods in the column names.
```{r}
library(stringr)
cn = colnames(circ)
cn = cn %>% 
  str_replace("Board", ".Board") %>% 
  str_replace("Alight", ".Alight") %>% 
  str_replace("Average", ".Average") 
colnames(circ) = cn
cn
```


## Removing the daily ridership

We want to look at each ridership, and will remove the `daily` column:
```{r}
circ$daily = NULL
```

```{r, echo = FALSE}
ex_wide = data.frame(id = 1:2,
                     visit1 = c(10, 5),
                     visit2 = c(4, 6),
                     visit3 = c(3, NA)
                     )
ex_long = data.frame(id = c(rep(1, 3), rep(2, 2)),
                     visit = c(1:3, 1:2),
                     value = c(10, 4, 3, 5, 6))
```

## Reshaping data from wide (fat) to long (tall)

See http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/

- Wide - multiple columns per observation 
    - e.g. visit1, visit2, visit3
```{r, echo = FALSE}
ex_wide
```

- Long - multiple rows per observation
```{r, echo = FALSE}
ex_long
```



## Reshaping data from wide (fat) to long (tall): base R

The `reshape` command exists.  It is a **confusing** function.  Don't use it.

## Reshaping data from wide (fat) to long (tall): tidyr {.smaller}

`tidyr::gather` - puts column data into rows.

We want the column names into "`var`" variable in the output dataset and the value in "`number`" variable. We then describe which columns we want to "gather:" 
```{r}
library(tidyr)
long = gather(circ, key = "var", value = "number", 
              starts_with("orange"),
              starts_with("purple"), 
              starts_with("green"),
              starts_with("banner"))
head(long, 2)
table(long$var)
```

## Reshaping data from wide (fat) to long (tall): tidyr

Now each `var` is boardings, averages, or alightings.  We want to separate these so we can have these by line.

```{r}
long = separate_(long, "var", 
                 into = c("line", "type"), 
                 sep = "[.]")
head(long, 3)
unique(long$line)
unique(long$type)
```

## Finding the First (or Last) record


```{r}
long = long %>% filter(!is.na(number) & number > 0)
first_and_last = long %>% arrange(date) %>% # arrange by date
  filter(type %in% "Boardings") %>% # keep boardings only
  group_by(line) %>% # group by line
  slice( c(1, n())) # select ("slice") first and last (n() command) lines
first_and_last %>%  head(4)
```

## Reshaping data from long (tall) to wide (fat): tidyr

In `tidyr`, the `spread` function spreads rows into columns.  Now we have a long data set, but we want to separate the Average, Alightings and Boardings into different columns:

```{r}
# have to remove missing days
wide = filter(long, !is.na(date))
wide = spread(wide, type, number)
head(wide)
```


## Reshaping data from long (tall) to wide (fat): tidyr

We can use `rowSums` to see if any values in the row is `NA` and keep if the row, which is a combination of date and line type has any non-missing data.

```{r}
# wide = wide %>%
#     select(Alightings, Average, Boardings) %>%
#     mutate(good = rowSums(is.na(.)) > 0)
namat = !is.na(select(wide, Alightings, Average, Boardings))
head(namat)
wide$good = rowSums(namat) > 0
head(wide, 3)
```

## Reshaping data from long (tall) to wide (fat): tidyr

Now we can filter only the good rows and delete the `good` column.
```{r}
wide = filter(wide, good) %>% select(-good)
head(wide)
```


## Data Merging/Append in Base R

* Merging - joining data sets together - usually on key variables, usually "id"
* `merge()` is the most common way to do this with data sets
* `rbind`/`cbind` - row/column bind, respectively
    * `rbind` is the equivalent of "appending" in Stata or "setting" in SAS
    * `cbind` allows you to add columns in addition to the previous ways
* `t()` is a function that will transpose the data

## Merging 

```{r merging}
base <- data.frame(id = 1:10, Age= seq(55,60, length=10))
base[1:2,]
visits <- data.frame(id = rep(1:8, 3), visit= rep(1:3, 8),
                    Outcome = seq(10,50, length=24))
visits[1:2,]
```

## Merging 

```{r merging2}
merged.data <- merge(base, visits, by="id")
merged.data[1:5,]
dim(merged.data)
```


## Merging 

```{r mergeall}
all.data <- merge(base, visits, by="id", all=TRUE)
tail(all.data)
dim(all.data)
```

## Joining in `dplyr`

* `?join` - see different types of joining for `dplyr`
* Let's look at https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

## Left Join

```{r left_join}
lj = left_join(base, visits)
dim(lj)
tail(lj)
```

## Right Join

```{r right_join}
rj = right_join(base, visits)
dim(rj)
tail(rj)
```

## Full Join

```{r full_join}
fj = full_join(base, visits)
dim(fj)
tail(fj)
```


## Perform Operations By Groups: dplyr

`group_by` is a form of replacement for `tapply` (not a complete replacement). 


We will use `group_by` to group the data by line, then use `summarize` (or `summarise`) to get the mean Average ridership:
```{r}
gb = group_by(wide, line)
summarize(gb, mean_avg = mean(Average))
```

## Perform Operations By Groups: dplyr with piping

Using piping, this is:

```{r}
wide %>% 
  group_by(line) %>%
  summarise(mean_avg = mean(Average))
```


## Perform Operations By Multiple Groups: dplyr 

This can easily be extended using `group_by` with multiple groups.  Let's define the year of riding:

```{r}
wide = wide %>% mutate(year = year(date),
                       month = month(date))
wide %>% 
  group_by(line, year) %>%
  summarise(mean_avg = mean(Average))
```

# Bonus slides - explore after visualization! 

## Perform Operations By Multiple Groups: dplyr {.smaller}

We can then easily plot each day over time:

```{r}
library(ggplot2)
ggplot(aes(x = date, y = Average, 
               colour = line), data = wide) + geom_line()
```

## Perform Operations By Multiple Groups: dplyr {.smaller}

Let's create the middle of the month (the 15th for example), and name it
mon.

```{r}
mon = wide %>% 
  dplyr::group_by(line, month, year) %>%
  dplyr::summarise(mean_avg = mean(Average))
mon = mutate(mon, 
             mid_month = dmy(paste0("15-", month, "-", year)))
head(mon)
```


## Perform Operations By Multiple Groups: dplyr {.smaller}

We can then easily plot the mean of each month to see a smoother output:

```{r}
ggplot(aes(x = mid_month,
               y = mean_avg, 
               colour = line), data = mon) + geom_line()
```



## Bonus! Points with a smoother! {.smaller}
```{r}
ggplot(aes(x = date, y = Average, colour = line), 
           data = wide) + geom_smooth(se = FALSE) + 
  geom_point(size = .5)
```




# Extra `group_by` examples

## `group_by` 

`group_by` is a form of replacement for `tapply` (not a complete replacement).

Example using Bike Lanes:
http://www.aejaffe.com/summerR_2016/data/Bike_Lanes.csv

```{r}
bike = read.csv(
  "http://www.aejaffe.com/summerR_2016/data/Bike_Lanes.csv",
  as.is = TRUE)
```

## Summarizing data with `group_by` and `summarize`


Average bike length BY project:

```{r, message = FALSE}
bike %>% 
  group_by(project) %>% 
  summarise(mean(length)) # get the average length
```

## Naming columns in output in `summarize`

Using `summarise/summarize(my_new_column_name = output )` allows you to name the column in the output: 

```{r, message = FALSE}
bike %>% 
  group_by(project) %>% 
  summarize(mean_length = mean(length)) %>% 
  head(4) # head ONLY for slide printing
```
